Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2021-09-14 vertical-thermal-review.Rmd
tweak isoprene paragraph

Biochemical protection against foliage light and heat damage also increases with irradiance, and thus tends to be higher in the upper canopy than in the understory (Table `r table_leaf_traits`). 
Carotenoids and xanthophyll cycle pigments -- violaxanthin, antheraxanthin and zeaxanthin (VAZ) -- can play a role in antioxidant scavenging and converting excess excitation energy into heat to reduce photoinhibition [@niinemetsPhotosynthesisResourceDistribution2007a; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998]. 
The dissipation of excess light energy is also important for leaf acclimation to higher temperatures if high $T_{leaf}$ impairs the photochemical energy dissipation pathway [@havauxTemperaturedependentAdjustmentThermal1996]. 
Heat-sensitive emissions of isoprene gas from tree leaves enhance photosynthetic thermal tolerance by regulating antioxidant defenses and other metabolic processes [@sharkeyIsopreneEmissionPlants2008; @taylorCapacityEmitIsoprene2019; @monsonLeafIsopreneEmission2021].

@@

Isoprene gas emission from tree leaves is associated with photosynthetic thermal tolerance [@taylorCapacityEmitIsoprene2019; @sharkeyIsopreneEmissionPlants2008], and within species, emission rates tend to increase toward brighter and hotter microenvironments [@niinemetsEmissionFactorVolatile2010]. 
Across landscapes, emitting species increase in relative abundance toward hotter climates, exceeding half of trees in warm tropical forests [@taylorIsopreneEmissionStructures2018a].
In contrast, vertical structuring of leaf isoprene emissions within forests may indicate the importance of temporal variability in thermal conditions as distinct from the long-term average.
A recent inventory in an Amazon forest showed that both the abundance of emitting species and average emission rates were highest in the mid-canopy region instead of the hotter and brighter upper canopy [@taylorNewFieldInstrument2021]. 
This observation is consistent with the hypothesis that temperature sensitive isoprene emissions enable real-time acclimation to rapidly changing leaf thermal environments typical of the mid-canopy region [@sharkeyIsopreneEmissionPlants2008], due to the sun's passage across a dynamically structured canopy surface. 
This review focuses primarily on variation in mean conditions across the vertical profile, but future work should seek to understand how temporal dynamics of leaf heating [see @leighInfluenceLeafSize2017] vary through the canopy and influence leaf function.

>>>>

Biochemical protection against foliage light and heat damage also increases with irradiance, and thus tends to be higher in the upper canopy than in the understory (Table `r table_leaf_traits`). 
Carotenoids and xanthophyll cycle pigments -- violaxanthin, antheraxanthin and zeaxanthin (VAZ) -- can play a role in antioxidant scavenging and converting excess excitation energy into heat to reduce photoinhibition [@niinemetsPhotosynthesisResourceDistribution2007a; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998]. 
The dissipation of excess light energy is also important for leaf acclimation to higher temperatures if high $T_{leaf}$ impairs the photochemical energy dissipation pathway [@havauxTemperaturedependentAdjustmentThermal1996]. 
Capacity for heat-sensitive emissions of isoprene gas from tree leaves enhances photosynthetic thermal tolerance by regulating antioxidant defenses and other metabolic processes [@sharkeyIsopreneEmissionPlants2008; @taylorCapacityEmitIsoprene2019; @monsonLeafIsopreneEmission2021].

@@

Emission of the VOC isoprene gas from tree leaves is associated with photosynthetic thermal tolerance [@taylorCapacityEmitIsoprene2019; @sharkeyIsopreneEmissionPlants2008], and within species, emission rates tend to increase toward brighter and hotter microenvironments [@niinemetsEmissionFactorVolatile2010]. 
Across landscapes, emitting species increase in relative abundance toward hotter climates, exceeding half of trees in warm tropical forests [@taylorIsopreneEmissionStructures2018a].
In contrast, vertical structuring of leaf isoprene emissions within forests may indicate the importance of temporal variability in thermal conditions as distinct from the long-term average.
A recent inventory in an Amazon forest showed that both the abundance of emitting species and average emission rates were highest in the mid-canopy region instead of the hotter and brighter upper canopy [@taylorNewFieldInstrument2021]. 
This observation is consistent with the hypothesis that temperature sensitive isoprene emissions enable real-time acclimation to rapidly changing leaf thermal environments typical of the mid-canopy region [@sharkeyIsopreneEmissionPlants2008], due to the sun's passage across a dynamically structured canopy surface. 
Future work seeking to understand how temporal dynamics of leaf heating [see @leighInfluenceLeafSize2017] vary through the canopy and influence leaf function will be important for resolving the role of isoprenes in forest thermal sensitivity across the vertical gradient.
